---
title: "STAT 536 - Case Study 4"
author: "Sam Lee, Gavin Hatch"
format: pdf
subtitle: "Targeted Marketing"
geometry: 
  - top=1in
  - left=1in
  - right=1in
  - bottom=1in
fontsize: 10pt
abstract: ""
header-includes:
  - \usepackage{amsmath}
  - \usepackage{fancyhdr}
  - \usepackage{threeparttable}
  - \pagestyle{fancy}
  - \fancyfoot[C]{Sam Lee and Gavin Hatch, MS in Statistics at Brigham Young University}
  - \fancyfoot[R]{\thepage}
  - \fancyhead{}
  - \renewcommand{\headrulewidth}{0pt}
  - \renewcommand{\footrulewidth}{0pt}
---

```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(knitr)
library(patchwork)
library(kableExtra)
library(pROC)

set.seed(536)

source("setup.R")
source("eda.R")
source("lasso.R")
source("probit.R")
source("pcr.R")
source("evaluate.R")
```

## Model Evaluation

We first evaluate our models using the in-sample and out-of-sample metrics of the $F_1$ Score and Area under the ROC Curve (\emph{AUC}).

```{r echo=F}
in_sample %>% kable()

out_of_sample %>% kable()
```

```{r echo=F, fig.cap="Most influential factors determined by PCR: As can be seen above, we classify the first partial component as \\emph{Marketing Interaction History}, the second partial component as \\emph{Customer Demographics}, and the third partial component as \\emph{Seasonal Timing}.", fig.width=18}
#| label: fig-pcr-loadings

visualize.loadings()
#ggsave("loadings.png", width=18, height=5)
```

We classify the impact on the probability of opening an account through partial component analysis as seen in @fig-pcr-loadings. The first partial component captures the impact of past marketing interactions on customer receptiveness. This suggests that customers with favorable prior campaign outcomes are more likely to engage with the possibility of opening an account. We also point out the bisection here between the \emph{timing} of when a customer was last contacted and whether they were contacted at all. Contacting customers is especially important, although the chances of a customer opening an account increases the more time has passed after they were contacted.

```{=tex}
\begin{table}[h!]
\centering
\begin{threeparttable}
\caption{Estimates and 95\% Confidence Intervals for Principal Component Regression}
\label{tab:pcr_confint}
\begin{tabular}{lcc}
    \toprule
    & \textbf{Estimate (50\%)} & \textbf{95\% Confidence Interval (2.5\%, 97.5\%)} \\
    \midrule
    \textbf{PC1} \\
    I(1 - contacted) *           & -0.1580 & (-0.1826, -0.1422) \\
    poutcomesuccess *            &  0.1501 & ( 0.1346,  0.1726) \\
    previous *                   &  0.1420 & ( 0.1256,  0.1647) \\
    I(pdays * (contacted)) *     &  0.1406 & ( 0.1263,  0.1617) \\
    poutcomenonexistent *        & -0.1316 & (-0.1538, -0.1175) \\
    \midrule
    \textbf{PC2} \\
    maritalsingle                &  0.0127 & ( 0.0013,  0.0249) \\
    maritalmarried *             & -0.0119 & (-0.0228, -0.0016) \\
    age                          &  0.0131 & ( 0.0023,  0.0234) \\
    jobblue-collar *             & -0.0291 & (-0.0365, -0.0211) \\
    educationuniversity.degree * &  0.0199 & ( 0.0132,  0.0265) \\
    \midrule
    \textbf{PC3} \\
    monthaug                     &  0.0060 & (-0.0003,  0.0142) \\
    age                          &  0.0131 & ( 0.0023,  0.0234) \\
    monthmay *                   & -0.0241 & (-0.0314, -0.0172) \\
    maritalsingle                &  0.0127 & ( 0.0013,  0.0249) \\
    educationuniversity.degree * &  0.0199 & ( 0.0132,  0.0265) \\
    \midrule
    \textbf{Intercept} \\
    Intercept                    & -0.0001 & (-0.0030,  0.0021) \\
    \bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note:} We include the top five most significant factors for each partial component. Significant factors (where the 95\% CI does not include zero) are marked with an asterisk (*). Confidence interval was calculated using $B=100$ \# of bootstrap samples. Estimates are given the interpretation of its effect on the log-odds of whether a customer opens an account. Covariates are \emph{scaled}.
\end{tablenotes}
\end{threeparttable}
\end{table}
```

```{r echo=F, fig.width=12, message=F, warning=F, fig.cap="Model predictions with respect to contact strategy. The figure on the left was fit with a GAM smoother."}
#| label: fig-contact

visualize.contacts()+visualize.social.media()
ggsave("contacting.png", width=12, height=5)
```

@fig-contact depicts the effect of contacting the campaigns have had on the probability of whether a customer has opened up a new account. We note that, in general, the more excessive contacting that takes place, the lower resulting expected probability that a customer will open up a new account. Despite the large spread, on average, social media contacts result in higher success rates than direct contacts.
